Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
miscellaneous edits

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Inducing Domain-Specific Languages for Bayesian Program Learning}

@@

\twocolumn[
\icmltitle{Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning}

@@

\begin{abstract}
  Successful approaches to program induction require a
  hand-engineered domain-specific language (DSL),
  constraining the space of allowed programs and imparting
  prior knowledge of the domain.
  We contribute a program learning algorithm
  called \system
  that infers a DSL
  and trains a neural network to
  efficiently search for programs in the learned DSL.
  We use our model to build circuits,
  edit strings, do symbolic regression,
  and synthesize functions on lists,
  in each case showing that
  \system learns a domain-specific vocabulary for expressing solutions to problems in the domain.

@@

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain-Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain-specific
knowledge about the space of programs. In
practice
we are often given only a few examples for each
program to be induced, and thus success often hinges on having a good
DSL to
provide a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered DSLs?

We propose \emph{learning} the DSL. We consider the setting where we
have a collection of related programming tasks, each specified by a
set of input/output examples. We do not assume that the tasks are
annotated with ground-truth programs. We typically will not try to
learn a DSL completely from scratch, but rather to start from a weaker
or more general set of primitives and
construct a richer, more powerful, and better-tuned DSL.

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL. The learned DSL
distills commonalities across programs that solve tasks, helping
the agent solve related program learning problems. The neural
network ensures that searching for program solutions remains tractable
even as the DSL (and hence the search space for programs) expands.

@@

Viewed as an inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference} for the generative %ritchie2016deep
model.

@@

  \node at (nn) {
    \begin{tikzpicture}[x=2.5cm,y=1.25cm,transform canvas={scale=0.15,shift={+(-1,2.5)}}]
      \tikzstyle{neuron}=[circle,fill=blue!50,minimum size=20pt]
      \fill[fill=white] (-0.25,-0.5) rectangle (2.25,-4.5);
      \node[rectangle] at (1,1) {};
      \foreach \name / \y in {1,...,4}
          \node[neuron] (I-\name) at (0,-\y) {};
      \foreach \name / \y in {1,...,3}
          \node[neuron] (H-\name) at (1,-\y-0.5) {};
      \foreach \name / \y in {1,...,4}
          \node[neuron] (O-\name) at (2,-\y) {};
      \foreach \source in {1,...,4}
          \foreach \dest in {1,...,3}
              \draw [-latex] (I-\source) -- (H-\dest);
      \foreach \source in {1,...,3}
          \foreach \dest in {1,...,4}
              \draw [-latex] (H-\source) -- (O-\dest);
    \end{tikzpicture}
  };
  \node[shift={+(0,-0.55)}] at (nn) { $q(x)$ };

@@

  \end{tikzpicture}
  \caption{(a) DSL $\mathcal{D}$ generates programs $p$ by sampling DSL primitives according to probabilities $\theta$. From each program we observe a task $x$ (program inputs/outputs). (b)  Neural network $q(\cdot )$, the \emph{recognition model}, regresses from $x$ to the parameters of a distribution over programs,
    $\theta^{(x)}$. Black/red arrows correspond to the generative/recognition models.}\label{graphicalModel}
\end{figure}

To perform inference, we take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a \textbf{Wake} cycle means using the recognition model to search for programs that solve tasks.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
 called \textbf{Sleep-R}.

@@

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
 \begin{table}[b]
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\code{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \code{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \code{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists& \code{(map (lambda (x) (+ x k)) l)}\\
   &\hspace{0.5cm} \emph{(add k to every element of list l)}
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
 \end{table}

@@

Our goal  is to induce a DSL while finding programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs; (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation is that these steps bootstrap off each other (Fig.~\ref{feeding}):
\\\noindent \textbf{Wake: Searching for programs.}  Our program search  is informed by both the DSL and the recognition model.
  When these improve, we find better programs.
\\\noindent\textbf{Sleep-R: Learning a recognition model.} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce a DSL from the programs we found which solve the tasks.
  As we solve more tasks, we can hone in on richer DSLs that more closely match the domain.

@@

Sec.~\ref{mathematicalFraming} frames this 3-step procedure as
a means of maximizing a lower bound on the posterior probability of the DSL given the tasks.
Sec.~\ref{explorationSection} explains how we search for programs that solve tasks;
Sec.~\ref{recognitionSection} explains how we train a neural network to accelerate program search; and
Sec.~\ref{grammarInductionSection} explains how we induce a DSL from programs.

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed by Fig.~\ref{graphicalModel}(a). Writing $J$ for the joint probability of $(\mathcal{D},\theta)$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
  \nonumber
    J(\mathcal{D},\theta)&\triangleq \probability[\mathcal{D}]\probability[\theta]\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* &= \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta \\
  \nonumber
    \theta^*& =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
The above equations summarize the problem from the point of view of an ideal Bayesian learner.
However, Eq.~\ref{intractableObjectives}
is wildly intractable because evaluating $J(\mathcal{D},\theta)$ involves
summing over the  infinite set of all possible programs.
In practice we will only ever be able to sum over a finite set of programs.
So, for each task, we define a finite set of programs, called a \emph{frontier}, and only marginalize over the frontiers:
\\\noindent\textbf{Definition.} A \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is a finite set of programs where $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

Using the frontiers we  define the following intuitive lower bound on the joint probability, which we call $\lowerBound$:
\begin{align}
 J\geq \lowerBound\triangleq\probability[\mathcal{D}]\probability[\theta]\prod_{x\in X} \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{align}

@@

%% In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
%% Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
%% \begin{align}
%%   A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
%% &+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
%%   \end{align}
\system does approximate MAP inference in the generative model of Fig.~\ref{graphicalModel}(a) by maximizing this lower bound on the joint probability,
alternating maximization w.r.t. the frontiers (Wake) and the DSL (Sleep-G):
%So interleaved with the DSL induction steps are program induction steps. Sec.~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $\lowerBound$ w.r.t. the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x,p|\mathcal{D},\theta]$
is large.
%% which we can accomplish by adding new programs to the frontiers means searching for new programs $p$ for task $x$
%% where  is large.
\\\noindent \textbf{DSL Induction: Maxing $\int \lowerBound\;\mathrm{d}\theta$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound\;\mathrm{d}\theta$.
Once we have a DSL $\mathcal{D}$ we can update $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.

@@

\textbf{Neural recognition model: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large, because including those programs
in the frontiers will most likely increase $\lowerBound$.  %% With $q$ in hand we can find programs for
%% the frontier $\mathcal{F}_x$ by searching for programs that maximize
%% $q(p|x)$.
Rather than directly predicting a distribution over $p$ conditioned on $x$,
the recognition model predicts a distribution $\theta^{(x)}$ over components of the DSL.
This taps into the intuition that programming is primarily a top-down activity:
as human programmers, we often first decide what kind of
programming constructs we might need to use,
and then we figure out how to assemble them in to the desired program.
%% This approach implements an amortized
%% inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
%% Fig.~\ref{graphicalModel}.

Our lower bound $\lowerBound$ is unconventional,
and one might wonder why we do not instead maximize an ELBO-style bound like in a VAE or in the EM algorithm.
We explain in the Supplement that maximizing an ELBO-style bound
leads to a pathological behavior that  causes the model to easily become trapped in local optima.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is
used in functional programming languages like OCaml.
Type variables are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes an input of type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \code{(lambda (x) x)}$:\alpha\to \alpha$.
We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
$p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
a type $\alpha$ with $\tau$ is to introduce constraints on the type
variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
See Supplement for more detail on program representation.
With this notation in hand we now define DSLs:

@@

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \code{(lambda (}var\code{) }body\code{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | p: \tau' \in \mathcal{D}\cup\mathcal{E}$ \\
     \hspace*{6.9em}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace*{8.8em}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k = 1}^K\gets\text{argTypes}(\tau')$ 
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \code{(}$e\;a_1\; a_2\; \cdots\; a_K$\code{)}
 \STATE\textbf{where:}
 \STATE \quad yield$(\tau) = \begin{cases}
    \text{yield}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
    \tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE \quad argTypes$(\tau) = \begin{cases}
    [\alpha] + \text{argTypes}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
    []   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}

>>>>

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Inducing Domain Specific Languages for Bayesian Program Learning}

@@

\twocolumn[
\icmltitle{Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning}

@@

\begin{abstract}
  Successful approaches to program induction require a
  hand-engineered domain-specific language (DSL),
  constraining the space of allowed programs and imparting
  prior knowledge of the domain.
  We contribute a program learning algorithm
  called \system
  that infers a DSL
  and trains a neural network to
  efficiently search for programs in the learned DSL.
  We use our model to build circuits,
  edit strings, do symbolic regression,
  and synthesize functions on lists,
  in each case showing that
  \system learns a domain-specific vocabulary for expressing solutions to the problems in the domain.

@@

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain
specific knowledge about the space of programs. In
practice
we are often given only a few examples for each
program to be induced, and thus success often hinges on having a good
DSL, to
provide a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered DSLs?

We propose \emph{learning} the DSL. We consider the setting where we
have a collection of related programming tasks, each specified by a
set of input/output examples. We do not assume that the tasks are
annotated with ground-truth programs. We typically will not try to
learn a DSL completely from scratch, but rather to start from a weaker
or more general set of primitives in a Lisp-like representation and
construct a richer, more powerful and better-tuned DSL.

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve each of the tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL. The DSL
distills commonalities across programs that solve the tasks, letting
the agent solve other related program learning problems. The neural
network ensures that searching for program solutions remains tractable
even as the DSL (and hence the search space for programs) expands.

@@

Viewed as an inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector, $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference} for the generative %ritchie2016deep
model.

@@

  \node at (nn) {
    \begin{tikzpicture}[x=2.5cm,y=1.25cm,transform canvas={scale=0.15,shift={+(-1,2.5)}}]
      \tikzstyle{neuron}=[circle,fill=blue!50,minimum size=20pt]
      \fill[fill=white] (-0.25,-0.5) rectangle (2.25,-4.5);
      \node[rectangle,width=2cm] at (1,1) {};
      \foreach \name / \y in {1,...,4}
          \node[neuron] (I-\name) at (0,-\y) {};
      \foreach \name / \y in {1,...,3}
          \node[neuron] (H-\name) at (1,-\y-0.5) {};
      \foreach \name / \y in {1,...,4}
          \node[neuron] (O-\name) at (2,-\y) {};
      \foreach \source in {1,...,4}
          \foreach \dest in {1,...,3}
              \draw [-latex] (I-\source) -- (H-\dest);
      \foreach \source in {1,...,3}
          \foreach \dest in {1,...,4}
              \draw [-latex] (H-\source) -- (O-\dest);
    \end{tikzpicture}
  };
  \node[shift={+(0,-0.55)}] at (nn) { $q(x)$ };

@@

  \end{tikzpicture}
  \caption{(a) DSL $\mathcal{D}$ generates programs $p$ by sampling DSL primitives with probabilities $\theta$. Observe tasks $x$ (program inputs/outputs). (b)  Neural network $q(\cdot )$, the \emph{recognition model}, regresses from $x$ to the parameters of a distribution over programs,
    $\theta^{(x)}$. Black/red arrows correspond to the generative/recognition models.}\label{graphicalModel}
\end{figure}

To perform inference, we take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a wake cycle means using the recognition model to search for programs that solve the tasks.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
 called \textbf{Sleep-R}.

@@

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
 \begin{table}
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\code{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \code{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \code{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists& \code{(map (lambda (x) (+ x a)) l)}\\
   &\hspace{0.5cm} \emph{(add a to every element of list l)}
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
 \end{table}

@@

Our goal  is to induce a DSL while finding programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs; (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation is that these steps bootstrap off each other (Fig.~\ref{feeding}):
\\\noindent \textbf{Wake: Searching for programs.}  Our program search  is informed by both the DSL and the recognition model.
When these improve we find better programs.
\\\noindent\textbf{Sleep-R: Learning a recognition model.} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.

@@

Sec.~\ref{mathematicalFraming} frames this 3-step procedure as
a means of maximizing a lower bound on the posterior probability of the DSL given the tasks.
Sec.~\ref{explorationSection} explains how we search for programs that solve the tasks;
Sec.~\ref{recognitionSection} how we train a neural network to accelerate the search over programs; and
Sec.~\ref{grammarInductionSection}  how \system induces a DSL from programs.

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Fig.~\ref{graphicalModel}(a). Writing $J$ for the joint probability of $(\mathcal{D},\theta)$, we want the $\mathcal{D}^*,\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)&\triangleq \probability[\mathcal{D}]P(\theta)\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* &= \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta \\
\nonumber  \theta^*& =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
The above equations summarize the problem from the point of view of an ideal Bayesian learner.
However, Eq.~\ref{intractableObjectives}
is wildly intractable because evaluating $J(\mathcal{D},\theta)$ involves
summing over the  infinite set of all possible programs.
In practice we will only ever be able to sum over a finite set of programs.
So, for each task, we define a finite set of programs, called a \emph{frontier}, and only marginalize over the frontiers:
\\\noindent\textbf{Definition.} A \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is a finite set of programs where $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

Using the frontiers we  define the following intuitive lower bound on the joint probability, which we call $\lowerBound$:
\begin{align}
 J\geq \lowerBound\triangleq\probability[\mathcal{D}]P(\theta)\prod_{x\in X} \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{align}

@@

%% In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
%% Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
%% \begin{align}
%%   A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
%% &+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
%%   \end{align}
\system does approximate MAP inference in the generative model of Fig.~\ref{graphicalModel}(a) by maximizing this lower bound on the joint probability,
alternatingly maximizing w.r.t. the frontiers (Wake) and the DSL (Sleep-G):
%So interleaved with the DSL induction steps are program induction steps. Sec.~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $\lowerBound$ w.r.t. the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x,p|\mathcal{D},\theta]$
is large.
%% which we can accomplish by adding new programs to the frontiers means searching for new programs $p$ for task $x$
%% where  is large.
\\\noindent \textbf{DSL Induction: Maxing $\int \lowerBound\;\mathrm{d}\theta$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound\;\mathrm{d}\theta$.
Once we have a DSL $\mathcal{D}$ we can update $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.

@@

\textbf{Neural recognition model: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large, because including those programs
in the frontiers will most increase $\lowerBound$.  %% With $q$ in hand we can find programs for
%% the frontier $\mathcal{F}_x$ by searching for programs that maximize
%% $q(p|x)$.
The recognition model exploits the structure of the DSL: %$\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a distribution over components of the DSL.
This taps into the intuition that programming is primarily a top-down activity:
as human programmers, we often first decide what kind of
programming constructs we might need to use,
and then we figure out how to assemble them in to the desired program.
%% This approach implements an amortized
%% inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
%% Fig.~\ref{graphicalModel}.

Our lower bound $\lowerBound$ is unconventional,
and one might wonder why we do not instead maximize an ELBO-style bound like in a VAE or in the EM algorithm.
We explain in the supplement that maximizing an ELBO-style bound
leads to a pathological behavior that  causes the model to easily become trapped in local optima.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is used in functional languages like OCaml.
Variables that range over types are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes as input type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \code{(lambda (x) x)}$:\alpha\to \alpha$.
We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
$p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
a type $\alpha$ with $\tau$ is to introduce constraints on the type
variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
See Supplement for more detail on program representation.
With this notation in hand we now define DSLs:

@@

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \code{(lambda (}var\code{) }body\code{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p: \tau' \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.1cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k = 1}^K\gets\text{argTypes}(\tau')$ 
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \code{(}$e\;a_1\; a_2\; \cdots\; a_K$\code{)}
 \STATE\textbf{where:}
 \STATE  yield$(\tau) = \begin{cases}
\text{yield}(\beta)   &\text{ if }$\tau = \alpha\to \beta$\\
\tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE\hspace{0cm} argTypes$(\tau) = \begin{cases}
[\alpha] + \text{argTypes}(\beta)   &\text{ if }$\tau = \alpha\to \beta$\\
[]   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:
(1) A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
(2) Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.
